Vragenlijst thesis evaluatie

### 2015-08-05 doc/template/chap-4.tex
Laatste hoofdstuk bijgewerkt

The  spreading of malware has already been extensively researched. Because of the many different types of propagation it is hard to define a single model that can model all of them. Modelling the spread of malware depends on two key factors: the method used for the propagation and the graph where the malware will spread itself. \\
Viruses and worms are the types of malware that are the most researched. Since the spreading of viruses requires human interaction, their propagation delay is depending on (hard to predict) human behaviour. Worms on the other hand, spread without human intervention, and their propagation is therefore easier to model. Since the purpose of this chapter is only to illustrate how a delay can be calculated, we will limit the presentation to propagation models of worms. \\

@@

Infection by worms start by dropping the worm on the network. Different dropping mechanisms can be used, whereby the initial attack can be random or targeted at a specific node. Frequently used mechanisms are USB sticks (given to a specific person or left behind for pick up by a random person), email, or malicious software through fishing or trojan horses. More important for determining our delay is the propagation strategy. Propagation is performed by determining the next nodes to spread the worm to. The following are common propagation methods: 
\begin{description}
%\item Scanning methods: A worm tries to guess (scan) the address of potential target nodes to infect. It can use distinct ways of scanning.
%% random, localized, topological or hit-list scanning.
%\begin{itemize}
\item \textbf{Selective Random scanning:} Some worms propagate by using the method of selective random targeting IP addresses. The worm selects randomly a part of the selected address space instead of scanning the whole address space. The reserved address blocks and the unassigned addresses are excluded from the address space. The rate of success for randomly chosen IP addresses is very low, but it is easy to implement. Example of such worms are `Code Red' and `Slammer'.
\item \textbf{Localized scanning:} A worm that uses localized scanning will scan for hosts in the local address space. This method is used by the `Code Red II' and `Nimda' worm.
\item \textbf{Topological scanning:} Address information stored in the victim machines are used to locate new targets. This method used by the `Morris' worm.
%\item \textbf{Hit-list scanning:} With the method of hit-list scanning, a short-list of vulnerable systems is made beforehand.  This is done to speed up the spread of worms at the initial stage. --This list consists of potentially vulnerable machines that are gathered beforehand and targeted first when the worm is released. An extreme case for the hit-list-scanning worms is a flash worm, which gathers all vulnerable machines into the list.--
\item \textbf{Sequential scanning:}
With sequential scanning, the worm will scan the IP addresses sequentially. This means that once a vulnerable host is compromised, it will look for IP addresses that are near to this host. For example, the address of the host is A, the next addresses that the worm will scan are A+1, or A-1. This method is used by the `Blaster worm'.
\item \textbf{DNS random scanning:} Another strategy is a kind of strategy in which the DNS infrastructure is used to locate a new target address. The IP address table from a DNS server is acquired from DNS records. The speed of a DNS scanning worm in the IPv6 internet is comparable to the speed of an IPv4 random scanning worm. There are some difficulties with this scanning technique. First it is difficult to obtain the whole address space from the DNS records. Secondly, the worm has to carry the database, which can slow down the propagation if the database is to big. Last, the IP addresses stored in the address table are only hosts with public domain names.
\item \textbf{Routable scanning:} The worms using routable scanning acquire target IP addresses based on the routing information in a network. Through  the BGP routing tables they can scan the routable address space. This method is three times faster than a traditional worm that uses random scanning. Examples are `Spyb0t' or `network Bluepill'. 
%--a ''routing worm'' [20]. Zou et al. designed two types of routing worms [20]. One type, based on Class-A (x.0.0.0/8) address allocations, is thus called 'ClassA routing worms.'' Such worms can reduce the scanning space to 45.3\% of the entire IPv4 address space. The other type, based on BGP routing tables, is thus called''BGP routing worms'' Such worms can reduce the scanning space to only about 28.6\% of the entire IPv4 address space.--
%\end{itemize}

\item \textbf{Topology-based Wroms} Email and other client application worms:An email worms uses the email systems to find email addresses to propagate. Other client applications can include: Internet Relay Chat (IRC), Instant Messenger (IM), and a variety of peer-to-peer
file sharing systems, which have been used by worms to propagate in a similar way as email worms. For example, the `Kak worm' is a Javascript computer worm that spread itself by exploiting a bug in Outlook Express. % Another example is `GhostNet'

@@

\subsection{Continious-time models}
\subsubsection{Simple Epidemic model}
%Codered paper gebruikt voor model
A Simple epidemic model is SI model. Nodes in the network can be either susceptible or infected.
Once a node is infected it will stay infected. Every node has the same
chance to be infected. The simple epidemic model is considered to be of a fixed size. The model for a fixed population is as follows:
\begin{equation}
\dfrac{d I(t)}{dt} = \beta I(t)[N-I(t)]
\end{equation} 
where I(t) is the number of infected hosts at time t; $\beta$ is the propagation rate; and N is the number of nodes in the network. In the beginning, $t=0$, I(0) hosts are infected. All the other nodes, $N - I(0)$, in the network are susceptible. 
The solution of this equation is the following logistic curve:
\begin{equation}
I = \dfrac{e^{\beta(t-T)}}{1+e^{\beta(t-T)}}
\end{equation}
where T is a time parameter representing the point of maximum increase in the
growth.\\

@@

Figuur \ref{SEMmodel} laat het SEM model zien.
This Simple epidemic model has been used in various papers \cite{OwnInternetSI}, \cite{CodeRed} to model random scanning worm such as Code Red and Slammer. \\
For our paper we have to approximate the propagation speed en Infect gelijkstellen aan het aantal genoeg nodes die geinfecteerd moeten zijn door de APT om de controle te hebben. Dan weten we hoelang het duurt. In het begin zullen er ook geen geinfecteerde hosts zijn, dus de vergelijking kan opgelost worden.\\
Er wordt geen rekening gehouden met de topology ? nog uitzoeken.
\subsubsection{Kermack-Mckendrick model: SIR}
In the Kermack-Mckendrick model, also known as SIR model, nodes can have three states: susceptible, infected or removed. Once a node of the network has been recovered from a worm, the node will stay in the removed mode and never becomes infected again. These nodes are not able to infect other nodes and can no longer be infected. \\
Let I(t) be the number of infectious hosts at time \textit{t}, R(t) be the number of removed hosts at time \textit{t} and J(t) is the number of infected hosts by time \textit{i}, regardless the fact that a node can be in a removed state.
\begin{equation}
J(t) = I(t) + R(t).
\end{equation} 
The Kermack-McKendrick model can be represented as follows:
\begin{equation}
\begin{Bmatrix} \dfrac{d J(t)}{dt} = \beta J(t) \big[N- J(t) \\
\dfrac{d R(t)}{dt} = \gamma I(t) \\
J(t) = I(t) + R(t) = N - S(t)
 \end{Bmatrix}
\end{equation}
Parameter $\beta$ is again the rate of infection and $\gamma$ is the rate of removal of infected hosts. S(T) is the number of susceptible hosts at time t.
\begin{figure}[hbtp]
\centering
\includegraphics[scale=0.5]{Images/KMmodel.png}
\caption{KM Model. The x-coordinate is the propagation time and the y-coordinate is the infected percentage of the whole network. Original image from \cite{OnWorms2005survey}. $N=10000, \beta = 1/10000000$.}
\label{KMmodel}
\end{figure}

%codered%
--Define $\rho = \gamma/\beta$ to be the relative removal rate [3]. One
interesting result coming out of this model is
dI(t)
dt > 0 if and only if S(t) > p. (6)
Since there is no new susceptible host to be generated,
the number of susceptible hosts S(t) is a monotonically decreasing
function of time t. If S(t0) < p, then S(t) < p and
dI(t)/dt < 0 for all future time t>t0. In other words, if the
initial number of susceptible hosts is smaller than some critical
value, S(0) < p, there will be no epidemic and outbreak
[15].
The Kermack-Mckendrick model improves the classical
simple epidemic model by considering that some infectious
hosts either recover or die after some time. However, this
model is still not suitable for modeling Internet worm propagation.
First, in the Internet, cleaning, patching, and filtering
countermeasures against worms will remove both susceptible
hosts and infectious hosts from circulation, but KermackMckendrick
model only accounts for the removal of infectious
hosts. Second, this model assumes the infection rate
to be constant, which isn't true for a rampantly spreading
Internet worm such as the Code Red worm--

%\subsubsection{Two-Factor model}
%%distributed worm simulation
%--Zou et
%al. [15] present a ''two-factor'' model that extends SIR
%epidemiological model to capture the effects of human
%countermeasures and the congestion due to the worm
%spread. Shen et al. [16] provide a discrete-time worm
%model that considers patching and cleaning effect and can
%model worms with local scanning techniques. All these
%approaches abstract specifics of the Internet topology,
%change in the size of vulnerable population as the worm
%spreads, and the effect of the individual host and network
%defenses (except for patching, in case of [16]) on the
%spread--
\subsubsection*{self disciplinary worms}
Model for self disciplinary worms and counter measures ... []

%Popular mechanism that worms use to detect vulnerable targets by random IP scanning probing. Feasible due to use 32-bit addresses. 128-bit addresses life harder for worms, except the ones that use email systems to propagate. two new strategies: uniformly distributed random number generator to select new target. :spread locally, by biasing the search space towards addresses within the same subnet or network.  
%The second strategy is almost the same as what email virusses would get. For this reason we work an example out .. 
%
%

\subsection{Continious-time models}
-- notities hier op overschrijven --
\subsubsection{OSN worms model}
\subsubsection{Email worms simulation model}

@@

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "thesis"
%%% End: 
\begin{figure}
\centering
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2.8cm,
                    semithick]
  \tikzstyle{every state}=[fill=blue,draw=none,text=white]

>>>>

The  spreading of malware has already been extensively researched. Because of the many different types of propagation it is hard to define a single model that can model all of them. Modelling the spread of malware depends on two key factors: the method used for the propagation and the graph of the network where the malware will spread itself. \\
Viruses and worms are the types of malware that are the most researched. Since the spreading of viruses requires human interaction, their propagation delay is depending on (hard to predict) human behaviour. Worms on the other hand, spread without human intervention, and their propagation is therefore easier to model. Since the purpose of this chapter is only to illustrate how a delay can be calculated, we will limit the presentation to propagation models of worms. \\

@@

Infection by worms start by dropping the worm on the network. Different dropping mechanisms can be used, whereby the initial attack can be random or targeted at a specific node. Frequently used mechanisms are USB sticks (given to a specific person or left behind for pick up by a random person), email, or malicious software through fishing or trojan horses. More important for determining our delay is the propagation strategy. Propagation is performed by determining the next nodes to spread the worm to. The following are common propagation methods: 
\begin{description}
%\item Scanning methods: A worm tries to guess (scan) the address of potential target nodes to infect. It can use distinct ways of scanning.
%% random, localized, topological or hit-list scanning.
%\begin{itemize}

\item \textbf{Selective Random scanning:} Some worms propagate by using the method of selective random targeting IP addresses. The worm selects randomly a part of the selected address space instead of scanning the whole address space. The reserved address blocks and the unassigned addresses are excluded from the address space. The rate of success for randomly chosen IP addresses is very low, but it is easy to implement. Example of such worms are \textit{Code Red} and \textit{Slammer}.

\item \textbf{Localized scanning:} A worm that uses localized scanning will scan for hosts in the local address space. This method is used by the \textit{Code Red II} and \textit{Nimda worm}.
%\item \textbf{Topological scanning:} Address information stored in the victim machines are used to locate new targets. This method used by the `Morris' worm.

%\item \textbf{Hit-list scanning:} With the method of hit-list scanning, a short-list of vulnerable systems is made beforehand.  This is done to speed up the spread of worms at the initial stage. --This list consists of potentially vulnerable machines that are gathered beforehand and targeted first when the worm is released. An extreme case for the hit-list-scanning worms is a flash worm, which gathers all vulnerable machines into the list.--

\item \textbf{Sequential scanning:}
With sequential scanning, the worm will scan the IP addresses sequentially. This means that once a vulnerable host is compromised, it will look for IP addresses that are near to this host. For example, the address of the host is A, the next addresses that the worm will scan are A+1, or A-1. This method is used by the \textit{Blaster worm}.

\item \textbf{DNS random scanning:} Another strategy is a kind of strategy in which the DNS infrastructure is used to locate a new target address. The IP address table from a DNS server is acquired from DNS records. 
The speed of a DNS scanning worm in the IPv6 internet is comparable to the speed of an IPv4 random scanning worm. 
%There are some difficulties with this scanning technique. 
%First it is difficult to obtain the whole address space from the DNS records. Secondly, the worm has to carry the database, which can slow down the propagation if the database is to big. Last, t
The IP addresses stored in the address table are only hosts with public domain names. This propagation method is used by \textit{MyDoom}.

\item \textbf{Routable scanning:} The worms using routable scanning acquire target IP addresses based on the routing information in a network. Through  the BGP routing tables they can scan the routable address space. This method is three times faster than a traditional worm that uses random scanning. Examples are \textit{Spyb0t} or \textit{network Bluepill}. 
%--a ''routing worm'' [20]. Zou et al. designed two types of routing worms [20]. One type, based on Class-A (x.0.0.0/8) address allocations, is thus called 'ClassA routing worms.'' Such worms can reduce the scanning space to 45.3\% of the entire IPv4 address space. The other type, based on BGP routing tables, is thus called''BGP routing worms'' Such worms can reduce the scanning space to only about 28.6\% of the entire IPv4 address space.--
%\end{itemize}

\item \textbf{Topology-based Worms} Email and other client application worms:An email worms uses the email systems to find email addresses to propagate. Other client applications can include: Internet Relay Chat (IRC), Instant Messenger (IM), and a variety of peer-to-peer
file sharing systems, which have been used by worms to propagate in a similar way as email worms. For example, the \textit{Kak worm} is a JavaScript computer worm that spread itself by exploiting a bug in Outlook Express. % Another example is `GhostNet'

@@

\subsection{Continious-time models}
\subsubsection{Simple Epidemic model}
%Codered paper gebruikt voor model
A Simple epidemic model is an SI model. Nodes in the network can be either susceptible or infected. 
Once a node is infected it will stay infected. Every node has the same chance to be infected
The simple epidemic model is considered to be of a fixed size. The model for a fixed population is as follows:

\begin{equation}
\dfrac{d I(t)}{dt} = \beta I(t)[N-I(t)]
\end{equation} 
where I(t) is the number of infected hosts at time t; $\beta$ is the propagation rate; and N is the number of nodes in the network. In the beginning, $t=0$, I(0) hosts are infected. All the other nodes, $N - I(0)$, in the network are susceptible. 
The solution of this equation is the following logistic curve:
\begin{equation}
I = \dfrac{e^{\beta(t-T)}}{1+e^{\beta(t-T)}}
\end{equation}
where T is a time parameter representing the point of maximum increase in the
growth.\\

@@

Figuur \ref{SEMmodel} laat het SEM model zien.
This Simple epidemic model has been used in various papers \cite{OwnInternetSI}, \cite{CodeRed} to model random scanning worm such as Code Red and Slammer. \\
For our paper we have to approximate the propagation speed en Infect gelijkstellen aan het aantal genoeg nodes die geinfecteerd moeten zijn door de APT om de controle te hebben. Dan weten we hoelang het duurt. In het begin zullen er ook geen geinfecteerde hosts zijn, dus de vergelijking kan opgelost worden.\\
%Er wordt geen rekening gehouden met de topology ? nog uitzoeken.

\subsubsection{RCS model}

RCS model stands for Random Constant Spread model and is developed by Paxson and Weaver at Stanford \citep{OwnInternetSI}. It is a model derived from the classical Simple Epidemic Model. They used this model to analyse the Code Red I worm. For this model it is assumed that the worm owns a good random number generator that is properly seeded. \\
Let N be the total of vulnerable hosts in the network that can be potentially infected. It is assumed that no system is patched, shut down, deployed or disconnected, which means that the number of hosts in the system stay constant. The model also ignores spreads of the worm behind firewalls on private networks. \\
Let K be the initial compromise rate. This is the rate of hosts that the worm can find and compromise per hour at the beginning of the infection. K is a global constant and therefore does not depend on the speed of the network or the processor speed.  Every machine can infect only one other machine after it has been compromised. It cannot increase the rate that a worm can find new hosts. 
The internet topology is considered as a complete undirected graph.  The model type is SI, so the state of the host can only be infected or suspected. Once the host is compromised it stays that way.  
Let T be the moment of the start of the infection. Variable \textit{a} is the proportion of vulnerable hosts that has been compromised. Variable \textit{t} is the time in hours. \\

The formula to model the spread of the worm is as follows:
\begin{equation}
N da = (N a)K(1 - a)dt
\end{equation}
It tells how many vulnerable machines will be compromised in the next amount of time \textit{dt}, when it is known the proportion of machines \text{a} that already have been compromised.\\

From this, it follows the simple differential equation:
\begin{equation}
\dfrac{d a}{dt} = Ka(1-a)
\end{equation}
With the following solution:
\begin{equation}
a = \dfrac{e^{K(t-T)}}{1+e^{K(t-T)}}
\end{equation}
\\

To to apply this formula for the case of FlipIt we need to know the maximum numbers of hosts that can be infected before the total system is compromised. The initial compromise rate has to be approximated. \textit{t} is the variable in our formula and says how many time has passed. If a passes a certain value we know the value of t which is the same as d in our FlipIt game.

%\subsubsection{Kermack-Mckendrick model: SIR}
%In the Kermack-Mckendrick model, also known as SIR model, nodes can have three states: susceptible, infected or removed. Once a node of the network has been recovered from a worm, the node will stay in the removed mode and never becomes infected again. These nodes are not able to infect other nodes and can no longer be infected. \\
%Let I(t) be the number of infectious hosts at time \textit{t}, R(t) be the number of removed hosts at time \textit{t} and J(t) is the number of infected hosts by time \textit{i}, regardless the fact that a node can be in a removed state.
%\begin{equation}
%J(t) = I(t) + R(t).
%\end{equation} 
%The Kermack-McKendrick model can be represented as follows:
%\begin{equation}
%\begin{Bmatrix} \dfrac{d J(t)}{dt} = \beta J(t) \big[N- J(t) \\
%\dfrac{d R(t)}{dt} = \gamma I(t) \\
%J(t) = I(t) + R(t) = N - S(t)
% \end{Bmatrix}
%\end{equation}
%Parameter $\beta$ is again the rate of infection and $\gamma$ is the rate of removal of infected hosts. S(T) is the number of susceptible hosts at time t.
%\begin{figure}[hbtp]
%\centering
%\includegraphics[scale=0.5]{Images/KMmodel.png}
%\caption{KM Model. The x-coordinate is the propagation time and the y-coordinate is the infected percentage of the whole network. Original image from \cite{OnWorms2005survey}. $N=10000, \beta = 1/10000000$.}
%\label{KMmodel}
%\end{figure}
%
%%codered%
%--Define $\rho = \gamma/\beta$ to be the relative removal rate [3]. One
%interesting result coming out of this model is
%dI(t)
%dt > 0 if and only if S(t) > p. (6)
%Since there is no new susceptible host to be generated,
%the number of susceptible hosts S(t) is a monotonically decreasing
%function of time t. If S(t0) < p, then S(t) < p and
%dI(t)/dt < 0 for all future time t>t0. In other words, if the
%initial number of susceptible hosts is smaller than some critical
%value, S(0) < p, there will be no epidemic and outbreak
%[15].
%The Kermack-Mckendrick model improves the classical
%simple epidemic model by considering that some infectious
%hosts either recover or die after some time. However, this
%model is still not suitable for modeling Internet worm propagation.
%First, in the Internet, cleaning, patching, and filtering
%countermeasures against worms will remove both susceptible
%hosts and infectious hosts from circulation, but KermackMckendrick
%model only accounts for the removal of infectious
%hosts. Second, this model assumes the infection rate
%to be constant, which isn't true for a rampantly spreading
%Internet worm such as the Code Red worm--
%
%%\subsubsection{Two-Factor model}
%%%distributed worm simulation
%%--Zou et
%%al. [15] present a ''two-factor'' model that extends SIR
%%epidemiological model to capture the effects of human
%%countermeasures and the congestion due to the worm
%%spread. Shen et al. [16] provide a discrete-time worm
%%model that considers patching and cleaning effect and can
%%model worms with local scanning techniques. All these
%%approaches abstract specifics of the Internet topology,
%%change in the size of vulnerable population as the worm
%%spreads, and the effect of the individual host and network
%%defenses (except for patching, in case of [16]) on the
%%spread--
\subsubsection*{self disciplinary worms}
Model for self disciplinary worms and counter measures ... []

%Popular mechanism that worms use to detect vulnerable targets by random IP scanning probing. Feasible due to use 32-bit addresses. 128-bit addresses life harder for worms, except the ones that use email systems to propagate. two new strategies: uniformly distributed random number generator to select new target. :spread locally, by biasing the search space towards addresses within the same subnet or network.  
%The second strategy is almost the same as what email virusses would get. For this reason we work an example out .. 
\subsection{Discrete-time models}
\subsubsection{Bluetooth worm model}
\todo{ref:Modeling Propagation Dynamics of Bluetooth Worms}

The Bluetooth worm model is introduced by Yan and Eidenbenz []. The model captures the behavior of the propagation of a worm that spreads through the Bluetooth protocol. 
A Bluetooth device that is compromised can only infect neighbour devices that are in its radio range
Let \textit{i(t)} be the average density of infected hosts in the network at time \textit{t}.

Formula is as follows: 
\begin{equation}
i(t_{k+1})=i(t_{k}) \cdot \dfrac{\rho(t_{k})}{i'(t_{k}) + (\rho(t_{k}) - i'(t_{k}))e^{-\alpha' \cdot \rho(t_{k}) / (\rho(t_{k}) - i'(t_{k}))}}
\end{equation}
where $\rho$(t) and $\beta$(t) are the average device density and the pairwise infection rate at time \textit{t} respectively. The number of new infections out of the infection cycle is denoted by $\alpha$(t).\\

The formula for a better estimation of the worm propagation
\begin{equation}
\alpha'=\dfrac{\rho(t_{k})-t(t_{k})}{\rho(t_{k})} \cdot \alpha(t_{k}) + \dfrac{i(t_{k})}{\rho(t_{k})} \cdot \alpha(t_{x})
\end{equation}

To apply this model on FlipIt we again set up a threshold value for i(t) and see how long it took to compromise this amount of hosts.  The paper concluded in their work  that after setting model parameters accordingly ($\lambda_{ne} = 0.2108$ the average node degree and $J_{in}= 0.2372$ the average meeting rate of neighbours), the model predicts that the time it would take to infect 99$\%$ of the devices is slightly less than one hour. So in this case the delay is equal to an hour and the amount of hosts that has to be infected by the attacker before the network is compromised is 99$\%$ of the hosts in the network.

\subsubsection{OSN worms model}
OSN worms stands for Online Social Network worms. This model is based on worms that propagate through the use of social media, like Facebook. The difference between an email worm and a social network worms is that the topology of the network is different.

%\subsubsection{AAWP model}
%SIR AAWP stands for Analytic Active Worm Propagation. The AAWP model is a discrete time propagation model. This model is used for worms that propagate through uniform scanning. When uniform scanning is used, each IP address 
%
%Popular mechanism that worms use to detect vulnerable targets by random IP scanning probing. Feasible due to use 32-bit addresses. 128-bit addresses life harder for worms, except the ones that use email systems to propagate. two new strategies: uniformly distributed random number generator to select new target. :spread locally, by biasing the search space towards addresses within the same subnet or network.  
%The second strategy is almost the same as what email virusses would get. For this reason we work an example out .. \\

A method to calculate the propagation of the virus in an easy way. Google page ranking algorithm.

@@

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "thesis"
%%% End: 
\begin{figure}
\centering
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2.8cm,
                    semithick]
  \tikzstyle{every state}=[fill=black!60!green,draw=none,text=white]
